Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures

### 2018-02-21 RNARedPrint-Bioinfo2018.tex

The \Def{dependency graph} induced by the set $\F$ of functions (on
sequences of length $n$) is the hypergraph
$G_\F=(\{1,\dots,n\} ,\{\dep(f)\mid f\in \F\})$.
Our algorithms will critically rely on a \Def{tree decomposition} of the dependency graph, which we define below.
\begin{definition}[Tree decomposition and width]
  \label{def:treedecomp}
  Given a hypergraph $G=(X, E)$, a \Def{tree decomposition} is a
  pair $(T,\chi)$, where $T$ is an unrooted tree/forest, and
  $\chi(v)\subseteq X$ is the set of vertices from $G$ assigned to the node $v\in T$, such
  that
\begin{enumerate}
\item each $x\in X$ occurs in at least one $\chi(v)$;
\item  $\forall x\in X$, $\{ v \mid x \in \chi(v) \}$ induces a connected subtree of $T$;
\item $\forall e\in E$ of $G$, there is a node $v\in T$, such that $e\subseteq\chi(v)$.
\end{enumerate}
The \Def{width} of a tree decomposition $(T,\chi)$ is defined as
$\width(T,\chi) = \min_{u\in T} |\chi(u)| - 1 $. The \Def{treewidth}
of $G$ is the smallest width in any tree decomposition of $G$.
\end{definition}
\section{Counting valid designs is \#{\sf P}-hard}\label{sec:counting}

%%% reformulate later
%We define the following notations: $\real^\infty:=\real\cup\{\infty\}$, $[i,j]:=\{i,\ldots,j\},$ \ldots
Now we turn to the complexity of $\NumDesign(G)$, the problem of computing the number of valid sequences for a given base pair dependency graph $G=(V,E)$. Note that this problem corresponds to the partition function problem in a simple $(0/\infty)$-valued base pair model, with $\beta\neq 1$. As previously noted~\citep{Flamm2001}, a set of target structures admits a valid design iff its base pair dependency graph is bipartite, which can be tested in linear time.
Moreover, without $\{\Gb, \Ub\}$ base pairs, any connected component  $C\in {\rm CC}(G)$ is entirely determined by the assignment of a single of its nucleotides. The number of valid designs is thus simply $4^{\#{\rm CC}(G)}$, where $\#{\rm CC}(G)$ is the \Def{number of connected components}.

  Let us now hypothesize the existence of a polynomial-time algorithm $\mathcal{A}$ for \NumDesign over strongly-connected graphs $G$. Now consider the polynomial-time algorithm $\mathcal{A}'$ that first executes $\mathcal{A}$ on $G$ to produce $\NumDesign(G)$, and returns $\NumDesign(G)/2=|\Design{G}| = |\IS{G}|$. We conclude that $\mathcal{A}'$ solves $\#{\sf BIS}$ in polynomial-time. This means that $\NumDesign$  is at least as hard as $\#{\sf BIS}$, thus does not admit a polynomial time exact algorithm unless $\#{\sf P}={\sf FP}$.
\end{proof}

@@

For our algorithmic description, we translate the concepts of
Section~\ref{sec:problem-statement} to the formalism of constraint networks, here
specialized as RNA design network. This allows us to base our
algorithm on cluster tree elimination (CTE) by Rina
Dechter~\citep{Dechter2013}.
%
In the RNA design network, (partially determined) RNA sequences
replace the more general concept of (partial) assignments in
constraint networks. Partially determined RNA sequences, called
\Def{partial sequences}, are words $\val$ over the alphabet
$\Sigma\cup\{?\}$ equivalently representing the set ${\mathcal S}(\val)$
of RNA sequences, where for positions $1\leq i\leq n$,
$\val_i\in\Sigma$ implies $S_i=\val_i$ for all $S\in{\mathcal
  S}(\val)$. The positions $1\leq i\leq n$, where $\val_i\in\Sigma$,
are called \Def{determined} by $\val$ and form its \Def{domain}.
%
Since the functions $f\in\F$ of Section~\ref{sec:problem-statement}
depend on only the subset $\dep(f)$ of sequence positions, one can
evaluate them for partial sequences $\val$ that determine (at least)
the nucleotides at all positions in $\dep(f)$. Thus for functions $f$
and partial sequences $\val$ that determine $\dep(f)$, we write
$\evalfor{f}{\val}$ to \Def{evaluate $f$ for $\val$}; i.e.{}
$\evalfor{f}{\val} := f(S),$ for any sequence
$S\in{\mathcal S}(\val)$.

@@

As already noted by \citet{Flamm2001}, the
requirement of canonical base pairing induces a bi-partition of each connected
component of the base pair dependency graph, such that the positions of one
partition are restricted to nucleotides $\{\Ab,\Gb\}$, but for other
partition, to nucleotides $\{\Cb,\Ub\}$. This suggests a strong
optimization of our algorithms, which reduces the basis in the
exponential factor from 4 to 2. The price for this reduction is a
binary choice for each of $c$ connected component. For each of the
$2^c$ combinations, the partition functions can be computed separately
in $O(n\,m\, 2^{w+1})$ time. This strategy is applied only if
$c< w+1$.
%% SW: This could be optimized further by applying the strategy only to (optimally) selected components
Finally, we compute the sums of the partition functions for all possible combinations.

\begin{theorem}[Complexities]
  For sequence length $n$, $k$ target structures, treewidth $w$ and
  a base pair dependency graph having $c$ connected components, $t$ sequences
  are sampled from the Boltzmann distribution  in
  %$O( n\, k \min(2^{w+c+1},4^{w+1}) + t\, n\, k )$ time.
  $\mathcal{O}( 2^d\, n\, k  + t\, n\, k )$ time, where $d:=\min(w+c+1,2(w+1))$.
\end{theorem}

@@

One then needs to \Def{learn a weights vector} $\pmb{\pi}$ such that, on average, the targeted energies are achieved by a random sequences in the weighted distribution, \emph{i.e.} such that  $\mathbb{E}(E_\ell(S)\mid \pmb{\pi})=\TargetE_\ell$,  $\forall\ell\in[1,k]$.
The expected value of $E_\ell$ is always decreasing for increasing weights $\pi_\ell$ (see Supp. Mat.~\ref{sec:weight_derivatives}). More generally, computing a suitable $\pmb{\pi}$ can be restated as a convex optimization problem, and be efficiently solved using a wide array of methods~\citep{Denise2010,Bendkowski2017}.
In practice, we use a simple heuristics which starts from an initial weight vector $\pmb{\pi}^{[0]}:=(e^\beta,\dots,e^\beta)$ and, at each iteration, generates a sample $\mathcal{S}$ of sequences. The expected value of an energy $E_\ell$ is estimated as $\hat\mu_\ell(\mathcal{S}) = \sum_{S\in\mathcal{S}}E_\ell(S)/|\mathcal{S}|$, and the weights are updated at the $t$-th iteration by %  $\pi_\ell^{[t+1]} = \pi_\ell^{[t]}\cdot\TargetE_\ell/\hat\mu_\ell(\mathcal{S})$.
$\pi_\ell^{[t+1]} = \pi_\ell^{[t]}\cdot \gamma^{\hat\mu_\ell(\mathcal{S})-\TargetE_\ell}$. In practice, the constant $\gamma>1$ is chosen empirically to achieve effective optimization.
While heuristic in nature, this basic iteration was elected in our initial version of \ourprog{} because of its reasonable empirical behavior (choosing $\gamma=1.2$).

@@

@@

\begin{table}
  \centering
  \caption{Trained weights for the base pair energy model.}
  \label{tab:basepairmodel}
  \begin{tabular}{c@{\quad}c@{\quad}c@{\quad}|@{\quad}c@{\quad}c@{\quad}c}
    \multicolumn{3}{c}{non-stacked} & \multicolumn{3}{c}{stacked}\\
    AU      & CG       & GU      & AU       & GC       & GU \\\hline
    1.26630 & -0.09070 & 0.78566 & -0.52309 & -2.10208 & -0.88474
  \end{tabular}
\end{table}

@@

\begin{table}
  \centering
  \caption{Trained weights for the stacking energy model (the rows specify $S_i,S_j$; the columns, $S_{i+1},S_{j-1}$; we do not show the symmetric weights).}
  \label{tab:stackingmodel}
  \begin{tabular}{c@{\quad}|@{\quad}c@{\quad}c@{\quad}c@{\quad}c@{\quad}c@{\quad}c@{\quad}c@{\quad}c}
       & AU & CG & GC & GU & UA & UG \\\hline
    AU &
-0.18826 &
-1.13291 &
-1.09787 &
-0.38606 &
-0.26510 &
-0.62086
    \\
    CG &
-1.11752 &
-2.23740 &
-1.89434 &
-1.22942 &
-1.10548 &
-1.44085
    \\
    GU &
-0.55066 &
-1.26209 &
-1.58478 &
-0.72185 &
-0.49625 &
-0.68876
    \\
  \end{tabular}
\end{table}

>>>>

The \Def{dependency graph} induced by the set $\F$ of functions (on
sequences of length $n$) is the hypergraph
$G_\F=(\{1,\dots,n\} ,\{\dep(f)\mid f\in \F\})$.
Our algorithms will critically rely on a \Def{tree decomposition} of the dependency graph, which we define below.
\begin{definition}[Tree decomposition and width]
  \label{def:treedecomp}
  Given a hypergraph $G=(X, E)$, a \Def{tree decomposition} is a
  pair $(T,\chi)$, where $T$ is an unrooted tree/forest, and
  $\chi(v)\subseteq X$ is the set of vertices from $G$ assigned to the node $v\in T$, such
  that
\begin{enumerate}
\item each $x\in X$ occurs in at least one $\chi(v)$;
\item  $\forall x\in X$, $\{ v \mid x \in \chi(v) \}$ induces a connected subtree of $T$;
\item $\forall e\in E$ of $G$, there is a node $v\in T$, such that $e\subseteq\chi(v)$.
\end{enumerate}
The \Def{width} of a tree decomposition $(T,\chi)$ is defined as
$\width(T,\chi) = \min_{u\in T} |\chi(u)| - 1 $. The \Def{treewidth}
of $G$ is the smallest width in any tree decomposition of $G$.
\end{definition}

@@

For our algorithmic description, we translate the concepts of
Section~\ref{sec:problem-statement} to the formalism of constraint networks, here
specialized as RNA design network. This allows us to base our
algorithm on the cluster tree elimination (CTE) of~\citet{Dechter2013}.
%
In the RNA design network, (partially determined) RNA sequences
replace the more general concept of (partial) assignments in
constraint networks. Partially determined RNA sequences, called
\Def{partial sequences}, are words $\val$ over the alphabet
$\Sigma\cup\{?\}$ equivalently representing the set ${\mathcal S}(\val)$
of RNA sequences, where for positions $1\leq i\leq n$,
$\val_i\in\Sigma$ implies $S_i=\val_i$ for all $S\in{\mathcal
  S}(\val)$. The positions $1\leq i\leq n$, where $\val_i\in\Sigma$,
are called \Def{determined} by $\val$ and form its \Def{domain}.
%
Since the functions $f\in\F$ of Section~\ref{sec:problem-statement}
depend on only the subset $\dep(f)$ of sequence positions, one can
evaluate them for partial sequences $\val$ that determine (at least)
the nucleotides at all positions in $\dep(f)$. Thus for functions $f$
and partial sequences $\val$ that determine $\dep(f)$, we write
$\evalfor{f}{\val}$ to \Def{evaluate $f$ for $\val$}; i.e.{}
$\evalfor{f}{\val} := f(S),$ for any sequence
$S\in{\mathcal S}(\val)$.

@@

\begin{theorem}[Complexities]\label{th:complexities}
  For sequence length $n$, $k$ target structures, treewidth $w$ and
  a base pair dependency graph having $c$ connected components, $t$ sequences
  are generated from the Boltzmann distribution  in
  $O( n\, k \, 4^{w+1} + t\, n\, k )$ time.
  %$O( n\, k \min(2^{w+c+1},4^{w+1}) + t\, n\, k )$ time.
  %$\mathcal{O}( 2^d\, n\, k  + t\, n\, k )$ time, where $d:=\min(w+c+1,2(w+1))$.
\end{theorem}
As shown in Supp. Mat.~\ref{sec:improvedComplexity}, the complexity of the precomputation can be further improved to 
$\mathcal{O}(n\,k\,2^{w+1}\,2^{c})$, where $c$ is the maximum number of connected components represented in a node of the tree decomposition ($c\le w+1$).

@@

One then needs to \Def{learn a weights vector} $\pmb{\pi}$ such that, on average, the targeted energies are achieved by a random sequences in the weighted distribution, \emph{i.e.} such that  $\mathbb{E}(E_\ell(S)\mid \pmb{\pi})=\TargetE_\ell$,  $\forall\ell\in[1,k]$.
The expected value of $E_\ell$ is always decreasing for increasing weights $\pi_\ell$ (see Supp. Mat.~\ref{sec:weight_derivatives}). More generally, computing a suitable $\pmb{\pi}$ can be restated as a convex optimization problem, and be efficiently solved using a wide array of methods~\citep{Denise2010,Bendkowski2017}.
In practice, we use a simple heuristics which starts from an initial weight vector $\pmb{\pi}^{[0]}:=(e^\beta,\dots,e^\beta)$ and, at each iteration, generates a sample $\mathcal{S}$ of sequences. The expected value of an energy $E_\ell$ is estimated as $\hat\mu_\ell(\mathcal{S}) = \sum_{S\in\mathcal{S}}E_\ell(S)/|\mathcal{S}|$, and the weights are updated at the $t$-th iteration by %  $\pi_\ell^{[t+1]} = \pi_\ell^{[t]}\cdot\TargetE_\ell/\hat\mu_\ell(\mathcal{S})$.
$\pi_\ell^{[t+1]} = \pi_\ell^{[t]}\cdot \gamma^{\hat\mu_\ell(\mathcal{S})-\TargetE_\ell}$. In practice, the constant $\gamma>1$ is chosen empirically to achieve effective optimization.
While heuristic in nature, this basic iteration was elected in our initial version of \ourprog{} because of its reasonable empirical behavior (choosing $\gamma=1.2$). Now, consider Given an elected vertex $v$ of a connected component, say the first node encountered in a rooted

@@

\section{Exploiting the bipartite nature of compatibility graphs to reduce the complexity}\label{sec:improvedComplexity}

As already noted by \citet{Flamm2001}, the
requirement of canonical base pairing induces a bi-partition of each connected
component within the compatibility graph, such that the nucleotides assigned to the two set of nodes in the partition are 
restricted to values $\{\Ab,\Gb\}$ and $\{\Cb,\Ub\}$ respectively. It follows that, within each connected component of the compatibility graph, the assignment of any position drastically restricts the domain of the remaining positions, \emph{i.e.} they may now be assigned to only 2 out of the original 4 nucleotides.
\newcommand{\Parity}[1]{p_{#1}}
To exploit this property, and further reduce the exponential factor of the complexity, we introduce the notion of \Def{parity} of a vertex/position, whereby vertices of \Def{even} (resp. \Def{odd}) polarities may only be assigned values in $\{\Ab,\Gb\}$ (resp. $\{\Cb,\Ub\}$). Neighbors of an odd vertex must be even and vice versa, so the choice of a parity for any single vertex caracterizes the parity of all vertices in a bipartite connected subgraph.

Let $CC=\{cc_1,cc_2,\ldots\}$ denote the connected components of the compatibility graph, we consider a modified version of the cluster tree, obtained by:
\begin{enumerate}
\item Electing, in each connected component $cc_i$, the first position $v_i^\star$ from the component encountered from the root in a pre-order traversal of the cluster tree;\label{step:election}
\item Introducing new variables $\mathcal{P}=\{\Parity{1},\ldots,\Parity{|CC|}\}$, representing the parity of elected vertices;
\item Adding, to each node $x$ in the tree decomposition, the parity variables of connected components that are represented in the node, \emph{i.e.} such that at least one of the positions in the connected component is in $\chi(x)$; \label{step:addparities}
\item Introducing, in the tree decomposition, a chain of intermediate nodes to fix cases where more than a single (parity) variable is introduced between a parent and its child.\label{step:intermediates}
\end{enumerate}
Note that, since connected components correspond to connected subtrees in the tree decomposition, then adding those variables preserves the properties of a tree decomposition. Also, each node in the former cluster tree introduces at most one new variable compared to its parent, so the elected node is uniquely defined by the rule stated in Step~\ref{step:election}. Finally, we note thatthe parity of a variable $v\in cc_i$ is entirely characterized by the parity $\Parity{i}$ of $v_i^\star$, and the parity of the distance between $v$ and $v_i^\star$ in the compatibility graph. Namely, the parity of $v$ is $\Parity{i}$ if the distance between $v$ and $v^{\star}_i$ is even, and opposite to $\Parity{i}$ if odd. It follows that, thanks to the new variables introduced in Step~\ref{step:addparities}, the domain of each of the original variables can be reduced to two alternatives instead of the original four.

The computation of Alg.~\ref{alg:pf} on the new cluster tree, considering reduced assignments for position variables and additional odd/even assignments for parity variables, has complexity in $\mathcal{O}(n\,k\,2^{w+1}\,2^{c})$, where $n$ is the RNA length, $k$ is the number of structures, $w$ is the tree-width of the unmodified cluster tree, and $c$ is the maximum number of connected components represented in a node. It should be noted that $c\le \min(|CC|,w+1)$, so that the new complexity cannot be larger than the one stated in Th.~\ref{th:complexities}. Also, note that the number of nodes introduced by Step~\ref{step:intermediates} is bounded by $|CC|$, so that the introduction of these intermediate nodes does not impact the worst-case complexity. For the same reason, the complexity of the stochastic backtrack remains unchanged.

@@

\begin{table}[b]
  \centering
  \caption{Trained weights for the base pair energy model.}
  \label{tab:basepairmodel}
  \begin{tabular}{c@{\quad}c@{\quad}c@{\quad}|@{\quad}c@{\quad}c@{\quad}c}
    \multicolumn{3}{c}{non-stacked} & \multicolumn{3}{c}{stacked}\\
    AU      & CG       & GU      & AU       & GC       & GU \\\hline
    1.26630 & -0.09070 & 0.78566 & -0.52309 & -2.10208 & -0.88474
  \end{tabular}
\end{table}

@@

\begin{table}[b]
  \centering
  \caption{Trained weights for the stacking energy model (the rows specify $S_i,S_j$; the columns, $S_{i+1},S_{j-1}$; we do not show the symmetric weights).}
  \label{tab:stackingmodel}
  \begin{tabular}{c@{\quad}|@{\quad}c@{\quad}c@{\quad}c@{\quad}c@{\quad}c@{\quad}c@{\quad}c@{\quad}c}
       & AU & CG & GC & GU & UA & UG \\\hline
    AU &
-0.18826 &
-1.13291 &
-1.09787 &
-0.38606 &
-0.26510 &
-0.62086
    \\
    CG &
-1.11752 &
-2.23740 &
-1.89434 &
-1.22942 &
-1.10548 &
-1.44085
    \\
    GU &
-0.55066 &
-1.26209 &
-1.58478 &
-0.72185 &
-0.49625 &
-0.68876
    \\
  \end{tabular}
\end{table}
